---
title: "Environmental Enteropathy: Bacterial Figures"
author: "Chandni Desai & Rachel Rodgers"
date: "July 30, 2019"
output: html_document
---

# Introduction

Code for generation of basic bacterial figures.

```{r load-data, message=FALSE}
load("../data/RDataObjects/Apr29.RData")

# Load libraries
library("phyloseq")
library("RColorBrewer")
library("vegan")
library("gridExtra")
library("knitr")
library("plotly")
library("ggpubr")
library("data.table")
library("microbiome")
library("DESeq2")
library("ggrepel")
library("pheatmap")
library("FSA")
library("gmodels")
library("multcomp")
library("writexl")
library("tidyverse")

source("./shared_R_scripts/Helper_Functions.R")

theme_set(theme_bw(base_size = 14))
```

# FIGURE 1: COHORT

## Figure 1 Panel A: Sampling timeline

```{r figure1pA-sampling-timeline}
# Add color to points based on disease status and line segements based on growth status (prior).
# Remove samples not having an EE status.
# Add growth status for patient 3 which was previously missing from metadata.
deltaHAZValues <- read.delim("../documents/analyzedSamplesWithDeltas.txt",
                             check.names = FALSE) # 84

patient3GrowthStatus <- deltaHAZValues %>% 
  filter(Subject_ID == 3) %>% 
  select(Manary_Barcode, Growth_Status_Prior) %>% 
  mutate_if(is.factor, as.character) %>% 
  deframe()

sd <- as(sd, "data.frame") %>% droplevels()

sd <- sd %>% 
  filter(!(is.na(EE_status_based_on_percL))) %>% 
  droplevels()

sd$X.SampleID <- as.character(sd$X.SampleID)
sd$Growth_Status_Prior <- as.character(sd$Growth_Status_Prior)

sd <- sd %>% 
  mutate(Growth_Status_Prior = ifelse(X.SampleID %in% names(patient3GrowthStatus),
                                      yes = patient3GrowthStatus[X.SampleID],
                                      no = Growth_Status_Prior))

sd$Growth_Status_Prior <- as.factor(sd$Growth_Status_Prior)
levels(sd$Growth_Status_Prior)

sd$Growth_Status_Prior <- factor(sd$Growth_Status_Prior,
                                 labels = c("Adequate", "none",  "Moderate", 
                                            "none", "Poor"))
sd$Growth_Status_Prior <- factor(sd$Growth_Status_Prior, 
                                 levels = c("Adequate", "Moderate", "Poor", "none"))

plot1_A <- ggplot(sd,
                  aes(x = age_months,
                      y = reorder(factor(patient_ID), age_months),
                      shape = EE_status_based_on_percL)) + 
  geom_line(inherit.aes = FALSE, 
            size = 1,
            aes(x = age_months,
                y = reorder(factor(patient_ID), age_months),
                group = patient_ID,
                color = Growth_Status_Prior)) +
  scale_color_manual(breaks = c("Adequate", "Moderate", "Poor"),
                                values = c("Red", "Blue", "Green", "Black")) +
  guides(shape = guide_legend(title = "EED Status"),
         color = guide_legend(title = "Growth Velocity")) +
  geom_point(size = 2.5) + 
  labs(y = "Patient ID", x = "Age in months") +
  theme(axis.title = element_text(size = 10),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 9))

set.seed(2137091)
plot1_A
```

## Figure 1 Panel B: perc L

```{r figure1pB-percent-L}
# points should be shaped to reflect their EE status
plot1_B <- ggplot(sd, aes(x = "", y = perc_L)) +
  geom_boxplot(outlier.shape = NA) +
  #geom_point(aes(shape = EE_status_based_on_percL)) +
  geom_jitter(alpha = 0.7, width = 0.15, size = 2.5,
              aes(shape = EE_status_based_on_percL)) +
  geom_hline(yintercept = 0.2, alpha = 0.5, linetype = 2) +
  geom_hline(yintercept = 0.45 ,alpha = 0.5, linetype = 2) +
  labs(y = "%L", x = "") +
  #guides(shape = guide_legend(title = "EE Status")) +
  theme(axis.title = element_text(size = 10),
        axis.text = element_text(size = 10),
        legend.position = "none")
set.seed(2137091)
plot1_B
```

## Figure 1 Panel C: delta HAZ

```{r figure1pC-delta-HAZ}
## Use data from file: plottingdeltaHAZ_dataforfigure.txt (delta HAZ for all samples)
deltaHAZ <- read.table(file = "../documents/plottingdeltaHAZ_dataforfigure.txt", 
                       header = TRUE)

# add Growth_Status_Prior
deltaHAZ <- deltaHAZ %>% 
  mutate("Growth_Status" = case_when(deltaHAZ > 0 ~ "Adequate",
                                     deltaHAZ >= -0.30 & deltaHAZ <= 0 ~ "Moderate",
                                     deltaHAZ < -0.30 ~ "Poor"))

deltaHAZ$Growth_Status <- factor(deltaHAZ$Growth_Status,
                                 levels = c("Adequate", "Moderate", "Poor"))

plot1_C <- ggplot(deltaHAZ, 
                  aes(x = "", y = deltaHAZ)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.7, width = 0.15, size = 2.5,
              aes(color = Growth_Status)) +
  scale_color_manual(values = c("Red", "Blue", "Green")) +
  geom_hline(yintercept = -0.3, alpha = 0.5, linetype = 2) +
  geom_hline(yintercept = 0,alpha = 0.5, linetype = 2) +
  labs(y = expression(Delta ~ HAZ), x = "") +
  #guides(color = guide_legend(title = "Growth Status")) +
  theme(axis.title = element_text(size = 10),
        axis.text = element_text(size = 10),
        legend.position = "none")
set.seed(2137091)
plot1_C
```

## ARRANGE Figure 1

```{r figure1-panel-arrange, fig.width=11, fig.height=8.5}
figure1 <- ggarrange(plot1_A, 
                     ggarrange(plot1_B, plot1_C,
                               labels = c("B", "C"),
                               heights = c(1, 1),
                               nrow = 2), 
                     ncol=2, 
                     labels = "A", 
                     widths = c(1,1),
                     common.legend = TRUE, legend = "bottom")
figure1
```

```{r figure-1-tiff, eval=FALSE}
tiff("../figures/second_rebuttal_figures/Figure1.tiff",
     units = "in", width = 11, height = 8.5, res = 300)
figure1
dev.off()


pdf("../figures/second_rebuttal_figures/Figure1.pdf", width = 11, height = 8.5)
figure1
dev.off()
```

# FIGURE 2: Bacterial Community Composition, Richness Regression, Alpha & Beta Diversity

## Figure 2 Panel A : Bacterial Community Composition

```{r figure2pA-bacterial-community-composition}
plot2_A <- ggplot(ps2.goodmoderatepoorgrowth.priorsamplesonly.phylum, 
                  aes(x = sample_name, 
                      y = Abundance, 
                      fill = Phylum)) + 
  geom_bar(stat = "identity", width = 1, position = "fill") +
  labs(y="Relative Abundance", x="") +
  facet_wrap(~Growth_Status_Prior, scales = "free_x") +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 10),
        strip.text = element_text(size = 10),
        legend.position = "bottom",
        legend.title = element_blank(),
        legend.text = element_text(size = 8),
        legend.box.margin = margin(-10, -10, -10, -10),
        legend.key.size = unit(0.4, "cm"),
        legend.spacing.x = unit(0.05, "cm"),
        panel.grid = element_blank()) +
  #guides(fill = guide_legend(override.aes = list(size = 1))) +
  scale_fill_brewer(palette="Set1")
plot2_A
```

## Fig 2 Panel B: Richness Regression with Age

```{r figure1pB-bacterial-richness-with-age}
## Use phyloseq object : ps2 (All Samples)

# Remove healthy samples from the data before plotting
sampleDataPS2 <- as(sample_data(ps2), "data.frame")

ps2EESamples <- subset_samples(ps2,
                               EE_status_based_on_percL != "No" &
                                 !(is.na(EE_status_based_on_percL))) # 2432 x 126

alphaDivPS2EESamples <- estimate_richness(ps2EESamples, measures = c("Observed", "Shannon"))

# merge with sample data
alphaDivPS2EESamples <- merge(sampleDataPS2, alphaDivPS2EESamples, 
                              by ="row.names", all = FALSE)

## Calculate regression coefficient
regression <- lm(Observed ~ age_months, data = alphaDivPS2EESamples)
summary(regression)

plot2_B <- ggplot(alphaDivPS2EESamples, aes(x = age_months, y = Observed)) + 
  geom_point(size = 1.5) + 
  stat_smooth(method = lm) +
  labs(y="Bacterial Richness", x = "Age in months") +
  annotate("text", x = 20, y = 325, 
           label = "paste(italic(R) ^ 2, \" = 0.346\")", parse = TRUE) +
  theme(axis.title = element_text(size = 10),
        axis.text = element_text(size = 10),
        panel.grid = element_blank())
set.seed(2137091)
plot2_B
```

## For Fig 2 Panels C & D:  Subset Alpha Div DF by Initial (Timepoint 1) and Midstudy (Timepoint 2)

```{r subset-alpha-div-df}
# timepoint 1 = "initial" stage
alphaDivT1 <- subset(ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria.rich,
                     ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria.rich$stage == "initial")

# timepoint 2 = "midstudy" stage
alphaDivT2 <- subset(ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria.rich,
                     ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria.rich$stage == "midstudy")

# for panels C, D
timepointLabels <- c("Growth Period 1", "Growth Period 2")
names(timepointLabels) <- c("initial", "midstudy")
```

## Figure 2 Panel C: Richness Boxplots by Growth Status

```{r figure2pC-bacterial-richness - paired}
#----- paired analysis -----#
plot2_C_paired <- ggplot(ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria.rich, 
                  aes(x = Growth_Status_Prior, y = Observed)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.7, width = 0.15, size = 1.5) +
  ylab("Richness") +
  stat_compare_means(label = "p.format", method = "kruskal.test", 
                     hide.ns = TRUE, 
                     label.x.npc = 0,
                     label.y.npc = 0.9,
                     size = 3.5) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 10),
        panel.grid = element_blank())

#----- unpaired analysis -----#
plot2_C_unpaired <- plot2_C_paired +
  facet_wrap(~ stage, labeller = labeller(stage = timepointLabels)) +
  theme(axis.title.y = element_blank())
set.seed(2908795)
plot2_C_unpaired
```

Dunn's post-hoc test for unpaired richness:

```{r richness-post-hoc}
dunnsRichnessT1 <- dunnTest(alphaDivT1$Observed, alphaDivT1$Growth_Status_Prior)
dunnsRichnessT1$res

dunnsRichnessT2 <- dunnTest(alphaDivT2$Observed, alphaDivT2$Growth_Status_Prior)
dunnsRichnessT2$res
```

## Figure 2 Panel D: Shannon Diversity Boxplots by Growth Status

```{r figure2pD-bacterial-shannon-diversity}
#----- paired analysis -----#
plot2_D_paired <- ggplot(ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria.rich, 
                  aes(x = Growth_Status_Prior, y = Shannon)) +
  geom_boxplot(outlier.shape = NA) +
  geom_jitter(alpha = 0.7, width = 0.15, size = 1.5) +
  ylab("Shannon Diversity") +
  stat_compare_means(label = "p.format", method = "kruskal.test", 
                     hide.ns = TRUE, 
                     label.x.npc = 0.0,
                     label.y.npc = 0.9,
                     size = 3.5) +
  theme(axis.title.x = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        axis.title.y = element_text(size = 10),
        panel.grid = element_blank())

#----- unpaired analysis -----#
plot2_D_unpaired <- plot2_D_paired +
  facet_wrap(~ stage, labeller = labeller(stage = timepointLabels)) +
  theme(axis.title.y = element_blank())
set.seed(2908795)
plot2_D_unpaired
```

Dunn's post-hoc test for unpaired diversity:

```{r diversity-post-hoc}
dunnsDiversityT1 <- dunnTest(alphaDivT1$Shannon, alphaDivT1$Growth_Status_Prior)
dunnsDiversityT1$res

dunnsDiversityT2 <- dunnTest(alphaDivT2$Shannon, alphaDivT2$Growth_Status_Prior)
dunnsDiversityT2$res
```

Wilcox p-values for alpha div by growth status:

```{r wilcox-on-alpha-by-growth-status}
growthCategories <- c("Adequate" = "Adequate", "Moderate" = "Moderate",
                      "Poor" = "Poor")

ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria.rich$Growth_Status_Prior <-
  as.character(ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria.rich$Growth_Status_Prior)


RunWilcox <- function(df, growthCat, alphaCat) {
  
  initialVec <- df %>% 
    dplyr::filter(Growth_Status_Prior == growthCat, stage == "initial") %>% 
    dplyr::pull(!!sym(alphaCat))
  
  midstudyVec <- df %>% 
    dplyr::filter(Growth_Status_Prior == growthCat, stage == "midstudy") %>% 
    dplyr::pull(!!sym(alphaCat))
    
  res <- wilcox.test(initialVec, midstudyVec)
  return(res)
}

wilcoxResultsRich <- map(.x = growthCategories,
                         .f = RunWilcox,
                         df = ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria.rich,
                         alphaCat = "Observed")
wilcoxResultsRich

wilcoxResultsDiv <- map(.x = growthCategories,
                        .f = RunWilcox,
                        df = ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria.rich,
                        alphaCat = "Shannon")
wilcoxResultsDiv
```

## Figure 2 Panel E: weighted Unifrac PCoA

```{r figure2pE-bacterial-pcoa-wUniFrac-paired, eval=FALSE}
#---- paired analysis -----#

# Extract eigenvalues from ordination object for easier customization of ggplot
eigenValues <- ord.ps2.goodmoderatepoor.wuni.prior$values$Relative_eig

plot2_EDF <- plot_ordination(ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria,
                             ord.ps2.goodmoderatepoor.wuni.prior,
                             justDF = TRUE)

plot2_E_paired <- ggplot(plot2_EDF,
                  aes_string(x = "Axis.1", y = "Axis.2",
                             color = "Growth_Status_Prior")) +
  xlab(paste0("PCoA 1 (", round(100*eigenValues[1], digits = 1), "%)")) +
  ylab(paste0("PCoA 2 (", round(100*eigenValues[2], digits = 1), "%)")) +
  geom_point(size=2, alpha = 0.7) +
  annotate("text", x = -0.08, y = 0.2, label = "p = 0.634", size = 3.5) +
  scale_color_manual(values = c("red", "blue", "green")) +
  theme(legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        legend.box.margin = margin(-10, -10, -10, -10),
        axis.title = element_text(size = 10))
plot2_E_paired
```

"Unpaired" weighted UniFrac Plots

```{r figure2pE-bacterial-pcoa-wUniFrac-unpaired}
#---- unpairedpaired analysis -----#

# subset physeq objects
psObjT1 <- subset_samples(ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria,
                          stage == "initial") # 43 samples

psObjT2 <- subset_samples(ps2.goodmoderatepoorgrowth.priorsamplesonly.bacteria,
                          stage == "midstudy") # 38 samples

# get wUniFrac ordination, extract eigen values
set.seed(21233406)
ordT1 <- ordinate(psObjT1, method = "PCoA", distance = "wUniFrac")
eigenValuesT1 <- ordT1$values$Relative_eig

set.seed(51773735)
ordT2 <- ordinate(psObjT2, method = "PCoA", distance = "wUniFrac")
eigenValuesT2 <- ordT2$values$Relative_eig

# ordination plot DFs
plot2_E_df_T1 <- plot_ordination(psObjT1, ordT1, justDF = TRUE)

plot2_E_df_T2 <- plot_ordination(psObjT2, ordT2, justDF = TRUE)

# plots
plot2_E_unpaired_T1 <- ggplot(plot2_E_df_T1,
                  aes_string(x = "Axis.1", y = "Axis.2", color = "Growth_Status_Prior")) +
  xlab(paste0("PCoA 1 (", round(100*eigenValuesT1[1], digits = 1), "%)")) +
  ylab(paste0("PCoA 2 (", round(100*eigenValuesT1[2], digits = 1), "%)")) +
  geom_point(size = 2, alpha = 0.7) +
  annotate("text", x = -0.05, y = 0.15, label = "p = 0.348", size = 3.5) +
  scale_color_manual(values = c("red", "blue", "green")) +
  ggtitle("Growth Period 1") +
  coord_fixed(sqrt(eigenValuesT1[2]/eigenValuesT1[1])) +
  theme(legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, size = 10),
        panel.grid = element_blank())
plot2_E_unpaired_T1

plot2_E_unpaired_T2 <- ggplot(plot2_E_df_T2,
                  aes_string(x = "Axis.1", y = "Axis.2", color = "Growth_Status_Prior")) +
  xlab(paste0("PCoA 1 (", round(100*eigenValuesT2[1], digits = 1), "%)")) +
  ylab(paste0("PCoA 2 (", round(100*eigenValuesT2[2], digits = 1), "%)")) +
  geom_point(size = 2, alpha = 0.7) +
  annotate("text", x = -0.25, y = 0.2, label = "p = 0.639", size = 3.5) +
  scale_color_manual(values = c("red", "blue", "green")) +
  ggtitle("Growth Period 2") +
  coord_fixed(sqrt(eigenValuesT2[2]/eigenValuesT2[1])) +
  theme(legend.position = "right",
        legend.title = element_blank(),
        legend.text = element_text(size = 10),
        axis.title = element_text(size = 10),
        plot.title = element_text(hjust = 0.5, size = 10),
        panel.grid = element_blank())
plot2_E_unpaired_T2
```

ADONIS p-values:

```{r get-adonis-values}
adonisValues <- map(.x = list("T1" = psObjT1, "T2" = psObjT2),
                    .f = RunAdonis,
                    category = "Growth_Status_Prior",
                    distance = "wUniFrac")

adonisValues
```

## ARRANGE Figure 2

Paired

```{r figure2-panel-arrange-paired, fig.width=8.5, fig.height=11, eval=FALSE}
fig2PairedTopRow <- ggarrange(plot2_A, plot2_B, labels = c("A", "B"), ncol = 2)

fig2PairedBottomRow <- ggarrange(plot2_C_paired, plot2_D_paired, plot2_E_paired,
                                 labels = c("C", "D", "E"), ncol = 3)

figure2Paired <- ggarrange(fig2PairedTopRow, fig2PairedBottomRow, nrow = 2)
figure2Paired
```

```{r eval=FALSE}
figure2 <- ggarrange(plot2_1, 
                     ggarrange(plot2_2, 
                               plot2_3, 
                               plot2_4, 
                               labels = c("B", "C", "D"), 
                               ncol=3, 
                               widths=c(0.2, 0.2, 0.3)), 
                     nrow = 2, 
                     labels = c("A"),
                     heights = c(1.1, 0.6, 0.75))
figure2
```

Unpaired

```{r figure2-panel-arrange-unpaired, fig.width=8.5, fig.height=11}
fig2UnpairedTopRow <- ggarrange(plot2_A, plot2_B, labels = c("A", "B"), ncol = 2)

richnessAnnotated <- annotate_figure(plot2_C_unpaired,
                                     left = text_grob("Richness", rot = 90,
                                                      size = 10))

diversityAnnotated <- annotate_figure(plot2_D_unpaired,
                                      left = text_grob("Shannon Diversity", rot = 90,
                                                       size = 10))

fig2UnpairedMiddleRow <- ggarrange(richnessAnnotated, 
                                   diversityAnnotated,
                                   ncol = 2,
                                   labels = c("C", "D"))

fig2UnpairedBottomRow <- ggarrange(plot2_E_unpaired_T1,
                                   plot2_E_unpaired_T2,
                                   labels = "E", ncol = 2,
                                   common.legend = TRUE,
                                   legend = "bottom")

figure2Unpaired <- ggarrange(fig2UnpairedTopRow, fig2UnpairedMiddleRow,
                             fig2UnpairedBottomRow, nrow = 3,
                             heights = c(0.75, 0.75, 1))
figure2Unpaired
```

```{r fig-2-tiff, eval=FALSE}
tiff("../figures/second_rebuttal_figures/Figure2.tiff", 
     units = "in", width = 8.5, height = 11, res = 300)
figure2Unpaired
dev.off()

pdf("../figures/second_rebuttal_figures/Figure2.pdf", width = 8.5, height = 11)
figure2Unpaired
dev.off()
```

# Figure 3: DESeq Results Heatmap

The DESeq model contains all three growth groups (adequate, moderate, poor).  We will generate a list of the unique differentially abundant taxa between the three pairwise comparisons: adequate vs. moderate, adequate vs. poor and moderate vs. poor, then make a heatmap displaying the significant taxa along the y-axis with adequate, moderate, and poor samples along the x (columns) and filled by mean normalized abundance.

```{r bacterial-DESeq-analysis}
sample_data(ps2.goodmoderatepoorgrowth.priorsamplesonly.prevfiltered)$patient_ID <- as.factor(sample_data(ps2.goodmoderatepoorgrowth.priorsamplesonly.prevfiltered)$patient_ID)

dds <- phyloseq_to_deseq2(ps2.goodmoderatepoorgrowth.priorsamplesonly.prevfiltered,
                          ~ bf_at_sampling + patient_ID + Growth_Status_Prior)

ddsAnalysis <- DESeq(dds)

ddsResAP <- results(ddsAnalysis, 
                    contrast = c("Growth_Status_Prior", "Adequate", "Poor"))

ddsResAM <- results(ddsAnalysis,
                    contrast = c("Growth_Status_Prior", "Adequate", "Moderate"))

ddsResMP <- results(ddsAnalysis,
                    contrast = c("Growth_Status_Prior", "Moderate", "Poor"))

deseqResultsList <- list("AP" = ddsResAP, "AM" = ddsResAM, "MP" = ddsResMP)

# For each data set, generate results table using 0.05 as the significance threshold
deseqResTableList <- pmap(.l = list(ddsResults = deseqResultsList),
                          .f = GenerateDESeqResultsTable,
                          physeq = ps2.goodmoderatepoorgrowth.priorsamplesonly.prevfiltered)

# From the deseqResTableList, extract significant results and apply a baseMean filter of 10
deseqSigTables <- map(.x = deseqResTableList,
                     .f = ~ dplyr::filter(.x, Significant == TRUE & baseMean > 10))

deseqSigOTUs <- map(.x = deseqSigTables,
                    .f = ~ dplyr::select(.x, OTU))

# Get all unique significant OTUs 
allSigOTUs <- Reduce(f = function(df1, df2) {merge(x = df1, y = df2,
                                                   by = "OTU", all = TRUE)},
                     x = deseqSigOTUs)

# Add taxonomy to allSigOTUs
taxTable <- data.frame(as(tax_table(ps2.goodmoderatepoorgrowth.priorsamplesonly.prevfiltered),
                          "matrix")) %>% 
  rownames_to_column(var = "OTU")

allSigOTUs <- merge(allSigOTUs, taxTable, by = "OTU", all = FALSE) 
```

```{r annotations-for-heatmap}
# row annotations (Genus Species)
allSigOTUs <- allSigOTUs %>% 
  dplyr::mutate("Taxon" = paste(Genus, Species),
                "Taxon" = str_replace_all(Taxon, "_", " "),
                "Taxon" = ifelse(Taxon == "NA NA",
                                 yes = paste("unclassified", Family, sep = " "),
                                 no = Taxon))

# Because there are some repeated Taxons (same genus species names) for distinct
#   OTUs, need to add numeric identifiers.  These need to be propogated to the 
#   Figure 3 heatmap, Figure S1 (correlation coefficient) heatmap, and DESeq2
#   results.
uniqueNamesSigOTUs <- allSigOTUs %>% 
  dplyr::select(OTU, Taxon) %>% 
  distinct()

taxonNameCounts <- uniqueNamesSigOTUs %>% 
  group_by(Taxon) %>% 
  summarize(count = n()) %>% 
  mutate("duplicated_taxon" = ifelse(count > 1, yes = TRUE, no = FALSE)) %>% 
  dplyr::select(-count) %>% 
  deframe()

# Add duplicated_taxon info to allSigOTUs
allSigOTUs <- allSigOTUs %>% 
  mutate("duplicated_taxon" = taxonNameCounts[Taxon],
         "taxon_number" = NA)

# Add numbers
dupeTaxonVec <- vector(mode = "character")

for (i in 1:nrow(allSigOTUs)) {
  
  currentDuplicated <- allSigOTUs[i, "duplicated_taxon"]
  
  if (currentDuplicated == TRUE) {
    currentTaxon <- allSigOTUs[i, "Taxon"]
    dupeTaxonVec <- c(dupeTaxonVec, currentTaxon)
    currentTaxonCount <- sum(dupeTaxonVec == currentTaxon)
    allSigOTUs[i, "taxon_number"] <- currentTaxonCount
  } else {
    allSigOTUs[i, "taxon_number"] <- NA
  }
  
}

# Now add numbers to taxon for those that are repeated
allSigOTUs <- allSigOTUs %>% 
  mutate("unique_taxon" = ifelse(duplicated_taxon == TRUE,
                                 yes = paste(Taxon, taxon_number, sep = " "),
                                 no = Taxon))

otuLUT <- allSigOTUs %>% 
  dplyr::select(OTU, unique_taxon) %>%
  deframe()

otuLevels <- c("unclassified Ruminococcaceae",
               "unclassified Lachnospiraceae",
               "Oscillibacter NA",
               "Clostridium sensu stricto NA 1", "Clostridium sensu stricto NA 2",
               "Clostridium sensu stricto NA 3",
               "Clostridium XlVa NA",
               "Enterococcus NA 1", "Enterococcus NA 2",
               "Megasphaera NA",
               "Megamonas funiformis",
               "Elusimicrobium NA",
               "Olsenella NA",
               "Bifidobacterium ruminantium",
               "Alloprevotella NA 1", "Alloprevotella NA 2",
               "Prevotella NA 1", "Prevotella NA 2", "Prevotella NA 3",
               "Prevotella NA 4", "Prevotella NA 5", "Prevotella NA 6",
               "Prevotella NA 7", "Prevotella NA 8", "Prevotella NA 9",
               "Bacteroides dorei",
               "Parabacteroides merdae",
               "Providencia NA",
               "Helicobacter fennelliae",
               "Treponema succinifaciens")
```

```{r extract-normalized-abundance-values}
# replace the counts in the physeq object so they're normalized like the dds object
replace_counts = function(physeq, dds) {
  dds_counts = DESeq2::counts(dds, normalized = TRUE)
  if (!identical(taxa_names(physeq), rownames(dds_counts))) {
    stop("OTU ids don't match")
  }
  otu_table(physeq) = otu_table(dds_counts, taxa_are_rows = TRUE)
  return(physeq)
}

physeqBacteriaNormalized <- replace_counts(physeq = ps2.goodmoderatepoorgrowth.priorsamplesonly.prevfiltered,
                                           dds = ddsAnalysis)

bacteriaNormalizedMelt <- psmelt(physeqBacteriaNormalized)

# filter for just the significant taxa, add unique_taxon name from otuLUT
bacteriaNormalizedMeltSigOTUs <- bacteriaNormalizedMelt %>% 
  filter(OTU %in% allSigOTUs$OTU) %>% 
  mutate(unique_taxon = otuLUT[OTU])

# add pseudocount
pseudoNorm <- bacteriaNormalizedMeltSigOTUs %>% 
  mutate(pseudo_norm = 1 + Abundance)

# average normalized OTU counts across growth categories
avgSigOTUNormCounts <- pseudoNorm %>%
  group_by(Growth_Status_Prior, unique_taxon) %>% 
  summarize(avg_norm_count = log2(mean(pseudo_norm)))

# pivot wider to get correct matrix form for heatmap (rows as taxa, columns as 
#   growth categories)
heatmapNormDF <- avgSigOTUNormCounts %>% 
  pivot_wider(names_from = Growth_Status_Prior,
              values_from = avg_norm_count)

heatmapNormDF$unique_taxon <- factor(heatmapNormDF$unique_taxon,
                                     levels = otuLevels) 

heatmapNormDF <- heatmapNormDF %>% 
  arrange(unique_taxon) %>%
  column_to_rownames(var = "unique_taxon")

heatmapNormMtx <- as.matrix(heatmapNormDF)

bacteriaHeatmap <- pheatmap(heatmapNormMtx, 
                            show_colnames = TRUE, 
                            show_rownames = TRUE,
                            cluster_cols = FALSE,
                            cluster_rows = FALSE,
                            cellwidth = 50)
```

```{r save-heatmap, eval=FALSE}
pdf("../figures/second_rebuttal_figures/Figure3_raw.pdf",
    height = 6, width = 8)
bacteriaHeatmap
dev.off()
```

```{r save-bacteriaNormalizedMeltSigOTUs}
# save bacteria normalized melt for correlation heatmap
saveRDS(bacteriaNormalizedMeltSigOTUs,
        "../data/RDataObjects/bacteriaNormalizedMeltSigOTUs.RDS")

# save otuLevels
saveRDS(otuLevels,
        "../data/RDataObjects/otuLevels.RDS")
```

```{r save-deseq-results-tables, eval=FALSE}
# add unique_taxon to the deseqSigTables
deseqSigTables <- map(.x = deseqSigTables,
                      .f = ~ .x %>% mutate("unique_taxon" = otuLUT[OTU]))

renamedSigTables <- deseqSigTables
names(renamedSigTables) <- c("bacterial_Adequate-Poor",
                             "bacterial_Adequate-Moderate",
                             "bacterial_Moderate-Poor")
write_xlsx(renamedSigTables,
           path = "../data/bacterial_DESeq_Tables.xlsx")
```

## FIGURE 7: BACTERIA-PHAGE ASSOCIATIONS/RELATIONSHIPS

## Part 1 : Relationship between Bacteria and Phage Richness 

```{r figure7p1}
plot7_1 <- ggplot(bacteria_phage_richnessdiversity, 
                  aes(Bacterial_Richness,y=Phage_Richness, color=Growth_Status_Prior)) +
  geom_point(aes(color=Growth_Status_Prior)) +
  stat_smooth(method = "lm", se=T) +
  labs(y="Phage Richness", x="Bacterial Richness",  color="Growth Velocity") +
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        #panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())

plot7_1
```

## Part 2 : Relationship between Bacteria and Phage Shannon Diversity

```{r figure2p2}
plot7_2 <- ggplot(bacteria_phage_richnessdiversity,
                  aes(Bacterial_ShannonDiversity,
                      y=Phage_ShannonDiversity,
                      color=Growth_Status_Prior)) +
  geom_point(aes(color=Growth_Status_Prior)) +
  stat_smooth(method = "lm", se=T) +
  labs(y="Phage Shannon Diversity", 
       x="Bacterial Shannon Diversity", 
       color="Growth Velocity") +
  theme(axis.title = element_text(size = 12),
        axis.text = element_text(size = 10),
        legend.title = element_text(size = 12),
        legend.text = element_text(size = 10),
        #panel.grid.minor = element_blank(),
        panel.grid.major = element_blank())
plot7_2
```

## ARRANGE Panel Figure 7

```{r figure7-panel-arrange, fig.width=7.5, fig.height=6}
figure7 <- ggarrange(plot7_1, plot7_2, nrow = 2,
                     labels = c("A", "B"))
figure7
```

## LM for Richness & Diversity by Growth Velocity Groups

```{r lm-richness}
growth_categories <- c("Adequate", "Moderate", "Poor")

## test ##
richnessDF <- bacteria_phage_richnessdiversity %>% 
  filter(Growth_Status_Prior == "Adequate")

lm_richness_adequate <- lm(Phage_Richness ~ Bacterial_Richness,
                           data = richnessDF)

# get adj r2
summary(lm_richness_adequate)$adj.r.squared

K <- diag(2)
lm_richness <- glht(lm_richness_adequate, linfct = K)
glsum <- summary(lm_richness)
summary(lm_richness)$test$pvalues
```

```{r figure-7-tiff, eval=FALSE}
tiff("../figures/Figure7.tiff",
     units = "in", width = 7, height = 6, res = 300)
figure7
dev.off()

pdf("../figures/Figure7.pdf", height = 6, width = 7)
figure7
dev.off()
```

## Supplemental Figure 2 : Correlation heatmap of significant bacteria vs phage abundances in samples from 2 growth velocities

```{r fig.width=7.5, fig.height=6, eval=FALSE}
CombinedHeatmaps <- readRDS("../data/RDataObjects/CombinedHeatmap.RDS")
oldHeatmapColNames <- colnames(CombinedHeatmaps)
newHeatmapColNames <- map_chr(.x = oldHeatmapColNames,
                              .f = ~ str_remove(., pattern = "\\.[:digit:]$"))
CombinedHeatmapsModified <- CombinedHeatmaps
colnames(CombinedHeatmapsModified) <- newHeatmapColNames

combinedHeatmapFigure <- pheatmap(CombinedHeatmapsModified,
                                  cluster_rows = FALSE,
                                  cluster_cols = FALSE)

```

```{r figure-S2-tiff, eval=FALSE}
tiff("../figures/FigureS2.tiff",
     units = "in", width = 7, height = 6, res = 300)
combinedHeatmapFigure
dev.off()
```

### Session Info

```{r session-info}
writeLines(capture.output(sessionInfo()), 
           "EE_Bacterial_Figures_session_info.txt")
Sys.Date()
getwd()
sessionInfo()
```



## Figure 5 B : Viral richness with age (move to phage figures)

```{r figure1p5-viral-richness-regression}
# import data from virus-side
fullViralSampleData <- readRDS("../data/RDataObjects/fullViralSampleData.RDS")
viralRichRegression <- lm(Observed ~ age_months, data = fullViralSampleData)
summary(viralRichRegression)
```

```{r figure1p5-viral-richness-with-age}
plot1_5 <- ggplot(fullViralSampleData, 
                  aes(x = age_months, y = Observed)) + 
  geom_point(size = 1.5) + 
  stat_smooth(method = lm) +
  labs(y="Viral Richness", x="Age in months") +
  annotate("text", x = 20, y = 600, label = "paste(italic(R) ^ 2, \" = 0.023\")", 
           parse = TRUE) +
  theme(axis.title = element_text(size = 10),
        axis.text = element_text(size = 10))
set.seed(2137091)
plot1_5
```