---
title: "EE Significant Bacteria-Phage Correlation Heatmaps"
author: "Rachel Rodgers"
date: "May 03, 2019"
output: html_document
---

```{r load-libraries}
library("pheatmap")
library("tidyverse")
```

```{r load-data}
# bacteria
bacteriaNormalizedMeltSigOTUs <- readRDS("../data/RDataObjects/bacteriaNormalizedMeltSigOTUs.RDS")
otuLevels <- readRDS("../data/RDataObjects/otuLevels.RDS") # for bacteria

# phage
phageNormalizedMeltSigOTUs <- readRDS("../data/RDataObjects/phageNormalizedMeltSigOTUs.RDS")
```

```{r sigBacteriaNlAbd}
# add more specific bacterial names
sigBacteriaNlAbd <- bacteriaNormalizedMeltSigOTUs %>% 
  dplyr::select(X.SampleID, Abundance, unique_taxon) %>% 
  dplyr::rename(unique_id = X.SampleID)

bacteriaNames <- unique(sigBacteriaNlAbd$unique_taxon)

length(bacteriaNames) # 30 taxa
```

```{r sigPhageNlAbd}
# add more specific phage names
sigPhageNlAbd <- phageNormalizedMeltSigOTUs %>% 
  dplyr::mutate("phage" = paste(Rank1, Rank3)) %>% 
  dplyr::select(unique_id, Abundance, phage)

phageNames <- unique(sigPhageNlAbd$phage)

length(unique(phageNames)) # 3 taxa
```

The data frames holding the normalized abundance values for significant taxa need to be converted from long to wide format.  Then the data frames need to be filtered down to just the samples that were analyzed and separated based on growth status prior to merging and running correlation tests.

```{r get-analyzed-sample-info}
# Only want to look at analyzed samples, read these in from all_analyzed_sample_data.txt
allAnalyzedSamples <- read.delim("../documents/all_analyzed_sample_data.txt",
                                 sep = "\t")
# get unique IDs of the analyzed samples we want for the correlation plots
allAnalyzedSamples <- allAnalyzedSamples %>% 
  mutate_if(is.factor, as.character) %>% 
  mutate("unique_id" = str_extract(string = sample_name,
                                   pattern = "NA[:digit:]+"))

analyzedSampleIDs <- pull(allAnalyzedSamples, unique_id) # 84 total

# make a look-up-table of samples (by their unique_id) and their growth status
#   so it can be added to the abundance data frames for later separation
#   of correlation plots by growth status prior
growthStatusLUT <- as.character(allAnalyzedSamples$Growth_Status_Prior)
names(growthStatusLUT) <- allAnalyzedSamples$unique_id
```

```{r sig-abd-long-to-wide}
# bacteria
sigBacteriaWide <- pivot_wider(sigBacteriaNlAbd,
                               names_from = "unique_taxon", 
                               values_from = "Abundance") # 81 samples

sigBacteriaWideFilt <- sigBacteriaWide %>% 
  filter(unique_id %in% analyzedSampleIDs) # 81 samples

# phage
sigPhageWide <- pivot_wider(sigPhageNlAbd,
                            names_from = "phage",
                            values_from = "Abundance") # 83 samples
  
sigPhageWideFilt <- sigPhageWide %>% 
  filter(unique_id %in% analyzedSampleIDs) # 83 samples
```

```{r merge-sig-abd-by-sample}
allSigTaxaAbdDF <- merge(sigBacteriaWideFilt, sigPhageWideFilt,
                         by = "unique_id", all = FALSE) # 80 samples

# add growth status column
allSigTaxaAbdDF <- allSigTaxaAbdDF %>% 
  mutate("Growth_Status_Prior" = growthStatusLUT[unique_id])
```

```{r test, eval=FALSE}
# filter by growth status
sigTaxaAbdPoor <- allSigTaxaAbdDF %>%
  filter(Growth_Status_Prior == "Poor") %>% 
  dplyr::select(-Growth_Status_Prior) %>% 
  column_to_rownames(var = "unique_id")

# run correlation
sigTaxaCorrPoor <- cor(sigTaxaAbdPoor, method = "spearman")

# get rid of redundant rows/columns
#   Remove bacteria from rows
#   Select bacteria from columns
sigTaxaCorrPoorReduced <- sigTaxaCorrPoor %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "row_taxa") %>% 
  filter(row_taxa %in% phageNames) %>% # only phage in rows
  dplyr::select(row_taxa, unname(bacteriaRenamingLUT)) %>% # unname keeps correct bacteria names in columns
  column_to_rownames(var = "row_taxa")

corrHeatmapPoor <- pheatmap(sigTaxaCorrPoorReduced,
                            na_col = "grey", cluster_rows = FALSE,
                            cluster_cols = FALSE)
```

```{r correlation-functions}
MakeCorrelationMatrix <- function(growthStatus) {

  # filter abundance data by growth status
  sigTaxaGrowthFiltered <- allSigTaxaAbdDF %>% 
    filter(Growth_Status_Prior == growthStatus) %>% 
    dplyr::select(-Growth_Status_Prior) %>% 
    column_to_rownames(var = "unique_id")
  
  # run correlation
  sigTaxaGrowthCorrelation <- cor(sigTaxaGrowthFiltered,
                                  method = "spearman")
  
  # get rid of redundant rows/columns in correlation matrix
  #   Remove bacteria from columns (keep phage in columns)
  #   Select bacteria from rows (keep phage in rows)
  sigTaxaCorrReduced <- sigTaxaGrowthCorrelation %>%
    as.data.frame() %>%
    rownames_to_column(var = "row_taxa") %>%
    filter(row_taxa %in% bacteriaNames) %>% # only bacteria in rows
    dplyr::select(row_taxa, phageNames) # only phage in columns
  # Add levels for bacteria (rows)
  sigTaxaCorrReduced$row_taxa <- factor(sigTaxaCorrReduced$row_taxa,
                                        levels = otuLevels)
  sigTaxaCorrReduced <- sigTaxaCorrReduced %>% 
    arrange(row_taxa) %>% 
    column_to_rownames(var = "row_taxa")
  
  return(sigTaxaCorrReduced)

}

PlotCorrelationHeatmap <- function(correlationMtx) {
  correlationHeatmap <- pheatmap(correlationMtx,
                                 na_col = "grey", cluster_rows = FALSE,
                                 cluster_cols = FALSE)
}
```

```{r plot-correlations}
correlationMatrices <- map(.x = list("Poor" = "Poor",
                                     "Moderate" = "Moderate",
                                     "Adequate" = "Adequate"),
                           .f = ~ MakeCorrelationMatrix(.x))

adequateHM <- pheatmap(correlationMatrices$Adequate,
         na_col = "grey", cluster_rows = FALSE, cluster_cols = FALSE)


moderateHM <- pheatmap(correlationMatrices$Moderate,
         na_col = "grey", cluster_rows = FALSE, cluster_cols = FALSE)

poorHM <- pheatmap(correlationMatrices$Poor,
         na_col = "grey", cluster_rows = FALSE, cluster_cols = FALSE)
```

Stick all the heatmaps together side-by-side:  Adequate-Moderate-Poor

```{r combined-heatmaps, fig.width=11, fig.height=7.5}
correlationMatricesArranged <- list("Adequate" = correlationMatrices$Adequate,
                                    "Moderate" = correlationMatrices$Moderate,
                                    "Poor" = correlationMatrices$Poor)

matricesCombined <- Reduce(f = function(df1, df2) {cbind(x = df1, y = df2)},
                           x = correlationMatricesArranged)

pheatmap(matricesCombined,
         na_col = "grey", cluster_rows = FALSE, cluster_cols = FALSE)

# fix column names
newColNames <- colnames(correlationMatrices$Adequate)
newColNamesFinal <- rep(newColNames, 3)

matricesCombinedFinal <- matricesCombined
colnames(matricesCombinedFinal) <- newColNamesFinal

finalHeatmap <- pheatmap(matricesCombinedFinal,
                         na_col = "grey", cluster_rows = FALSE, 
                         cluster_cols = FALSE)
```

```{r, eval=FALSE}
ggarrange(adequateHM$gtable,
          moderateHM$gtable,
          poorHM$gtable,
          nrow = 3)
```


```{r save-comined-cor-heatmap, eval=FALSE}
pdf("../figures/second_rebuttal_figures/FigureS1_option1.pdf",
    width = 8.5, height = 11)
finalHeatmap
dev.off()
```



 
