---
title: "Environmental Enteropathy: Viral Richness Over Time"
author: "Rachel Rodgers"
date: "July 30, 2019"
output: html_document
---

```{r initiate-environment, include=FALSE}
# Global knitr option
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE)

# Load libraries
library("phyloseq")
library("RColorBrewer")
library("vegan")
library("gridExtra")
library("knitr")
library("plotly")
library("ggpubr")
library("data.table")
library("microbiome")
library("DESeq2")
library("ggrepel")
library("pheatmap")
library("FSA")
library("gmodels")
library("tidyverse")

source("./shared_R_scripts/Helper_Functions.R")
```

In this markdown I will be combining the selected phage from MEGAN and the eukaryotic virus data (at family level) from VS for all samples (healthy and EE) into one object in order to plot richness of viral taxa by age for figure 1.

First I will load the phage physeq object (physeqSubsetFinal).

```{r phage-data}
load("../data/RDataObjects/EE_Virome_Preprocessing.RData")
physeqSubsetFinal # 1021 x 145
head(sample_names(physeqSubsetFinal)) # contins all samples (healthy + EE)
```

Now I will load in the eukaryotic viral data for all samples.

```{r physeqEukFamily}
physeqEukFamily <- readRDS("../data/physeqObjects/physeqEukFamily.RDS")
physeqEukFamily # 34 x 145
head(sample_names(physeqEukFamily)) # contains all samples (healthy + EE)
```

All we need is an OTU table in order to calculate richness.

```{r}
# extract and transform the otu tables for phage and eukaryotes
phageOTUTable <- as(otu_table(physeqSubsetFinal), "matrix") %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "long_sample_name") %>% 
  mutate("unique_id" = str_extract(long_sample_name, 
                                   pattern = "(-|_)[:digit:]+(-|_)LNS(-|_)?[:digit:]+(-|)"),
         "unique_id" = str_remove_all(unique_id, pattern = "(-|_)")) %>% 
  select(long_sample_name, unique_id, everything())
dim(phageOTUTable) # 145 samples x 1023 taxa

eukOTUTable <- as(otu_table(physeqEukFamily), "matrix") %>% 
  t() %>% 
  as.data.frame() %>% 
  rownames_to_column(var = "short_sample_name") %>% 
  rename(unique_id = short_sample_name)
dim(eukOTUTable) # 145 samples x 35 taxa

# merge phage and eukaryotic otu tables on unique_id
fullViralOTUTable <- merge(phageOTUTable, eukOTUTable,
                           by = "unique_id",
                           all = TRUE) # 145 x 1057

# change the table back into otu class
viralOTUTable <- fullViralOTUTable %>% 
  select(-long_sample_name) %>% 
  column_to_rownames(var = "unique_id") %>% 
  as.matrix() %>% 
  t()
dim(viralOTUTable) # 1055 taxa x 145 samples

viralOTUTableFinal <- otu_table(viralOTUTable, taxa_are_rows = TRUE)

# Calculate richness on this table
totalViralRichness <- estimate_richness(viralOTUTableFinal,
                                        measures = "Observed")
row.names(totalViralRichness) <- c(str_replace(string = rownames(totalViralRichness),
                                               pattern = "^X",
                                               replacement = ""))
totalViralRichness <- rownames_to_column(totalViralRichness, var = "unique_id")

# merge with sample data, expect 157 rows
# add an ID column for merging to sampleDataCorrected
sampleDataCorrectedWithID <- sampleDataCorrected %>% 
  rownames_to_column(var = "sample_long_name") %>% 
  mutate("id" = str_extract(sample_name,
                            pattern = "(-|_)[:digit:]+(-|_)LNS(-|_)?[:digit:]+(-|)"),
         "id" = str_remove_all(id, pattern = "(-|_)"))

fullViralSampleData <- merge(sampleDataCorrectedWithID, totalViralRichness,
                             by.x = "id", by.y = "unique_id", all = TRUE)

saveRDS(fullViralSampleData, file = "../data/RDataObjects/fullViralSampleData.RDS")
```

```{r regression-viral-richness-by-age}
viralRichRegression <- lm(Observed ~ age_months, data = fullViralSampleData)
summary(viralRichRegression)

viralRichPlot <- ggscatter(fullViralSampleData, x = "age_months", y = "Observed", 
                           add = "reg.line",
                           conf.int = TRUE) +
  stat_smooth(method = lm) +
  geom_point(size = 2) +
  ggtitle("Viral Richness with Age") +
  labs(y = "Richness", x = "Age in months") +
  theme(plot.title = element_text(hjust = 0.5))

viralRichPlot
```

Save fullViralSampleData for use in EE_Bacterial_Figures.Rmd (for figure panel 1).

```{r save-fullViralSampleData}
saveRDS(fullViralSampleData, "../data/RDataObjects/fullViralSampleData.RDS")
```

### Session Info

```{r session-info}
writeLines(capture.output(sessionInfo()), 
           "EE_Viral_Richness_Over_Time_session_info.txt")
Sys.Date()
getwd()
sessionInfo()
```
